---
title: "Cytokine 'deconvolution' analysis"
output: html_notebook
---
   
# Setup   
## Load packages   
```{r}
library(tidyverse)
library(ComplexHeatmap)
library(janitor)
library(viridis)
library(ggrepel)
library(ggsci)
library(Cairo)
library(circlize)
library(ggsignif)

pal = pal_npg("nrc")(9)
set.seed(666)
pal_long = sample(colorRampPalette(pal_npg("nrc")(9))(28))
set.seed(666)
pal_medium = sample(colorRampPalette(pal_npg("nrc")(9))(16))
```
   
## Import mean expression data   
### scRNA   
```{r}
gene_conv = data.frame(gene = c("CCL21", "CXCL13", "CCL27", "EGF", "CXCL5",
                                "CCL11", "CCL24", "CCL26", "FGF2", "FLT3LG",
                                "CX3CL1", "CSF3", "CSF2", "CXCL1", "CCL1",
                                "IFNA2", "IFNG", "IL10", "IL13", "IL15",
                                "IL16", "IL17A", "IL1A", "IL1B", "IL1RN",
                                "IL2", "IL20", "IL21", "IL23", "IFNL2",
                                "IL3", "IL33", "IL4", "IL5", "IL6",
                                "IL7", "IL8", "IL9", "CXCL10", "LIF",
                                "CCL2", "CCL8", "CCL7", "CCL13", "CCL22",
                                "CCL3", "CCL4", "CCL15", "CD40LG", "KITLG",
                                "CXCL12", "CCL17", "TGFA", "TNF", "LTA",
                                "THPO", "TNFSF10", "TSLP", "VEGFA"),
                       display_name = c("CCL21", "CXCL13", "CCL27", "EGF", "CXCL5", "CCL11",
                                        "CCL24", "CCL26", "FGF-2", "FLT-3L", "CX3CL1",
                                        "CSF3", "CSF2", "CXCL1", "CCL1", "IFNa2",
                                        "IFNg", "IL-10", "IL-13", "IL-15", "IL-16",
                                        "IL-17A", "IL-1a", "IL-1b", "IL1RN", "IL-2",
                                        "IL-20", "IL-21", "IL-23", "IFNL2", "IL-3",
                                        "IL-33", "IL-4", "IL-5", "IL-6", "IL-7",
                                        "IL-8", "IL-9", "CXCL10", "LIF", "CCL2",
                                        "CCL8", "CCL7", "CCL13", "CCL22", "CCL3",
                                        "CCL4", "CCL15", "sCD40L", "SCF", "CXCL12",
                                        "CCL17", "TGFa", "TNFa", "LTA", "TPO",
                                        "TRAIL", "TSLP", "VEGF-A"))

#NOTE: these CSVs are COVID, only!
celltypes = read.csv("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_celltype_expression.csv") %>%
  dplyr::rename(celltype = Celltype) %>% 
  dplyr::filter(celltype != "Mixed Myeloid") %>% #meaningless
  pivot_longer(cols = -celltype, names_to = "gene", values_to = "log_rna_expression")

typestates = read.csv("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_subcluster_expression.csv") %>%
  dplyr::rename(cluster = Cluster) %>% 
  dplyr::filter(cluster != "Mixed myeloid") %>% #meaningless
  pivot_longer(cols = -cluster, names_to = "gene", values_to = "log_rna_expression")
```   
   
Add undetected genes as zeros   
```{r}
gene_conv = data.frame(gene = c("CCL21", "CXCL13", "CCL27", "EGF", "CXCL5",
                                "CCL11", "CCL24", "CCL26", "FGF2", "FLT3LG",
                                "CX3CL1", "CSF3", "CSF2", "CXCL1", "CCL1",
                                "IFNA2", "IFNG", "IL10", "IL13", "IL15",
                                "IL16", "IL17A", "IL1A", "IL1B", "IL1RN",
                                "IL2", "IL20", "IL21", "IL23", "IFNL2",
                                "IL3", "IL33", "IL4", "IL5", "IL6",
                                "IL7", "IL8", "IL9", "CXCL10", "LIF",
                                "CCL2", "CCL8", "CCL7", "CCL13", "CCL22",
                                "CCL3", "CCL4", "CCL15", "CD40LG", "KITLG",
                                "CXCL12", "CCL17", "TGFA", "TNF", "LTA",
                                "THPO", "TNFSF10", "TSLP", "VEGFA"),
                       display_name = c("CCL21", "CXCL13", "CCL27", "EGF", "CXCL5", "CCL11",
                                        "CCL24", "CCL26", "FGF-2", "FLT-3L", "CX3CL1",
                                        "CSF3", "CSF2", "CXCL1", "CCL1", "IFNa2",
                                        "IFNg", "IL-10", "IL-13", "IL-15", "IL-16",
                                        "IL-17A", "IL-1a", "IL-1b", "IL1RN", "IL-2",
                                        "IL-20", "IL-21", "IL-23", "IFNL2", "IL-3",
                                        "IL-33", "IL-4", "IL-5", "IL-6", "IL-7",
                                        "IL-8", "IL-9", "CXCL10", "LIF", "CCL2",
                                        "CCL8", "CCL7", "CCL13", "CCL22", "CCL3",
                                        "CCL4", "CCL15", "sCD40L", "SCF", "CXCL12",
                                        "CCL17", "TGFa", "TNFa", "LTA", "TPO",
                                        "TRAIL", "TSLP", "VEGF-A"))

# generate dataframes with zeros for each filtered gene for all cell types, states
undetected_genes = gene_conv %>% 
  dplyr::filter(!(gene %in% celltypes$gene)) %>% 
  .$gene

# celltypes
all_celltypes = unique(celltypes$celltype)
celltypes_zeros = data.frame(celltype = rep(all_celltypes, each = length(undetected_genes)),
                             gene = rep(undetected_genes, length(all_celltypes)),
                             log_rna_expression = 0)
celltypes = bind_rows(celltypes, celltypes_zeros) %>% 
  left_join(., gene_conv)

# typestates
all_typestates = unique(typestates$cluster)
typestates_zeros = data.frame(cluster = rep(all_typestates, each = length(undetected_genes)),
                             gene = rep(undetected_genes, length(all_typestates)),
                             log_rna_expression = 0)
typestates = bind_rows(typestates, typestates_zeros) %>% 
  left_join(., gene_conv)
```
   
### Cytokines (summarized, COVID only)   
```{r}
covid_cytokines = readRDS("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230308_5PL_fit_data_complete.rds") %>% 
  dplyr::filter(diagnosis == "COVID-19") %>% 
  group_by(display_name) %>% 
  dplyr::summarize(protein_expression = mean(mean_concentration, na.rm = T),
                   log_protein_expression = log10(protein_expression))
```
   
### Import metadata conversion   
```{r}
script_md = readRDS("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230308_5PL_fit_data_complete.rds") %>% 
  dplyr::select(study_id, tc_pt_study_id, diagnosis, Superinfection) %>% 
  unique()
```

   
# Heatmaps of expression   
## Cell type   
```{r}
#generate scaled matrix
celltype_mat = celltypes %>% 
  pivot_wider(id_cols = display_name, names_from = celltype, values_from = log_rna_expression) %>% 
  column_to_rownames("display_name") %>% 
  data.matrix()

celltype_hm = 
  Heatmap(matrix = celltype_mat,
        name = "Log-Normalized Expression",
        cluster_rows = T,
        cluster_columns = T,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_column_names = T, 
        col = inferno(100),
        column_names_rot = 45)

CairoPDF("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_celltype_deconv_hm.pdf",
    width = 12,
    height = 10, 
    family = "Arial")
celltype_hm
dev.off()
CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_celltype_deconv_hm.png",
    width = 12,
    height = 10, 
    units = "in",
    res = 300,
    family = "Arial")
celltype_hm
dev.off()

celltype_hm
```
   
## Type and state   
```{r}
#generate scaled matrix
typestate_mat = typestates %>% 
  pivot_wider(id_cols = display_name, names_from = cluster, values_from = log_rna_expression) %>% 
  column_to_rownames("display_name") %>% 
  data.matrix()

typestate_hm = 
  Heatmap(matrix = typestate_mat,
        name = "Log-Normalized Expression",
        cluster_rows = T,
        cluster_columns = T,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_column_names = T, 
        col = inferno(100),
        column_names_rot = 45)

CairoPDF("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_typestate_deconv_hm.pdf",
    width = 12,
    height = 10, 
    family = "Arial")
typestate_hm
dev.off()
CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_typestate_deconv_hm.png",
    width = 12,
    height = 10, 
    units = "in",
    res = 300,
    family = "Arial")
typestate_hm
dev.off()

typestate_hm
```   
   
# Relative contributions   
## Bin by log expression   
   
### Determine cutoffs   
```{r}
range(celltypes$log_rna_expression)
range(typestates$log_rna_expression)
```   
   
0 --> No contribution   
(0 - 1] --> Minor contributor   
(1 - 2] --> Mmoderate contributor   
(2 - Inf] --> Major contributor   
   
### By cell type   
```{r}
celltype_contributions = celltypes %>% 
  dplyr::mutate(contribution = case_when(log_rna_expression == 0 ~ "No contribution",
                                         log_rna_expression > 0 & log_rna_expression <= 1 ~ "Minor contributor",
                                         log_rna_expression > 1 & log_rna_expression <= 2 ~ "Moderate contributor",
                                         log_rna_expression > 2 ~ "Major contributor"),
                rna_expression = 10 ^ log_rna_expression) %>% 
  left_join(., covid_cytokines)

celltype_contribution_scatterplot = celltype_contributions %>% 
  ggplot(aes(x = reorder(display_name, -protein_expression), y = rna_expression)) +
  geom_col(aes(y = protein_expression), 
           position = "dodge", alpha = 0.01) +
  geom_point(alpha = 0.7, aes(color = contribution)) +
  geom_text_repel(data = subset(celltype_contributions, log_rna_expression > 0), 
                  aes(label = celltype, color = contribution), 
                  seed = 12345, min.segment.length = 0) +
  theme_bw() +
  scale_color_npg() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(x = "Analyte", y = "Log-Normalized Expression") +
  scale_y_log10()

CairoPDF("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_celltype_contributions_scatter.pdf",
    width = 12,
    height = 8, 
    family = "Arial")
celltype_contribution_scatterplot
dev.off()
CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_celltype_contributions_scatter.png",
    width = 12,
    height = 8, 
    units = "in",
    res = 300,
    family = "Arial")
celltype_contribution_scatterplot
dev.off()

celltype_contribution_scatterplot
```
   
Ben's alt version (stacked bar plots for scRNA)   
```{r}
celltype_contribution_barplot = celltype_contributions %>% 
  group_by(display_name) %>% 
  dplyr::filter(mean(log_rna_expression, na.rm = T) > 0) %>% 
  dplyr::mutate(moam_contribution = cur_data()$rna_expression[cur_data()$celltype == "MoAM"] /
                  sum(rna_expression, na.rm = T)) %>% 
  ggplot(aes(x = reorder(display_name, -moam_contribution), 
             y = rna_expression, group = rna_expression)) +
  geom_col(aes(fill = celltype), position = "fill") +
  theme_bw() +
  scale_fill_manual(values = pal_medium) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(x = "Analyte", y = "Proportion of scRNA Counts")

CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221117_celltype_stacked_rna.png",
    width = 14,
    height = 8, 
    units = "in",
    res = 300,
    family = "Arial")
celltype_contribution_barplot
dev.off()

celltype_contribution_barplot
```
   
### By type and state   
```{r}
typestate_contributions = typestates %>% 
  dplyr::mutate(contribution = case_when(log_rna_expression == 0 ~ "No contribution",
                                         log_rna_expression > 0 & log_rna_expression <= 1 ~ "Minor contributor",
                                         log_rna_expression > 1 & log_rna_expression <= 2 ~ "Moderate contributor",
                                         log_rna_expression > 2 ~ "Major contributor"),
                rna_expression = 10 ^ log_rna_expression) %>% 
  left_join(., covid_cytokines)

typestate_contribution_scatterplot = typestate_contributions %>% 
  ggplot(aes(x = reorder(display_name, -protein_expression), y = rna_expression)) +
  geom_col(aes(y = protein_expression), 
           position = "dodge", alpha = 0.005) +
  geom_point(alpha = 0.7, aes(color = contribution)) +
  geom_text_repel(data = subset(typestate_contributions, log_rna_expression > 0), 
                  aes(label = cluster, color = contribution), 
                  seed = 12345, min.segment.length = 0) +
  theme_bw() +
  scale_color_npg() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(x = "Analyte", y = "Log-Normalized Expression") +
  scale_y_log10()

CairoPDF("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_typestate_contributions_scatter.pdf",
    width = 12,
    height = 8, 
    family = "Arial")
typestate_contribution_scatterplot
dev.off()
CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221111_typestate_contributions_scatter.png",
    width = 12,
    height = 8, 
    units = "in",
    res = 300,
    family = "Arial")
typestate_contribution_scatterplot
dev.off()

typestate_contribution_scatterplot
```   
   
Ben's alt version (stacked bar plots for scRNA)   
```{r}
typestate_contribution_barplot = typestate_contributions %>% 
  group_by(display_name) %>% 
  dplyr::filter(mean(log_rna_expression, na.rm = T) > 0) %>% 
  dplyr::mutate(moam2_contribution = sum(cur_data()$rna_expression[grepl("MoAM", cur_data()$cluster)], 
                                         na.rm = T) /
                  sum(rna_expression, na.rm = T)) %>% 
  ggplot(aes(x = reorder(display_name, -moam2_contribution), 
             y = rna_expression, group = rna_expression)) +
  geom_col(aes(fill = cluster), position = "fill") +
  theme_bw() +
  scale_fill_manual(values = pal_long) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(x = "Analyte", y = "Proportion of scRNA Counts")

CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221117_typestate_stacked_rna.png",
    width = 16,
    height = 8, 
    units = "in",
    res = 300,
    family = "Arial")
typestate_contribution_barplot
dev.off()

typestate_contribution_barplot
```

   
# Direct correlation   
## By celltype   
### Organize data   
```{r}
sample_conv = read_csv("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/script-to-sc-patient.csv",
                       col_names = c("tc_pt_study_id", "Patient"))
complete_cytokines = readRDS("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230308_5PL_fit_data_complete.rds") %>% 
  dplyr::filter(sample_origin == "BAL") %>% 
  dplyr::select(tc_pt_study_id, display_name, 
                protein_expression = mean_concentration, diagnosis) %>% 
  dplyr::mutate(protein_expression = protein_expression,
                log_protein_expression = log10(protein_expression + 1))

#includes non-COVID!
celltype_cor_data = read.csv("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_celltype_expression_patient.csv") %>% 
  dplyr::filter(Celltype != "Mixed Myeloid") %>% #meaningless
  left_join(., sample_conv) %>% 
  dplyr::select(-Patient) %>% 
  pivot_longer(cols = -c(Celltype, tc_pt_study_id),
               names_to = "gene",
               values_to = "log_rna_expression") %>% 
  dplyr::mutate(rna_expression = 10 ^ log_rna_expression) %>% 
  left_join(gene_conv) %>% 
  left_join(., complete_cytokines)
```   
   
### Run correlations   
```{r}
celltype_cors = celltype_cor_data %>% 
  group_by(display_name, Celltype) %>% 
  dplyr::summarize(pearson_r = tryCatch(cor.test(rna_expression, protein_expression,
                                                 method = "pearson")$estimate,
                                        error = function(cond){
                                          return(NA_real_) }),
                   pearson_pval = tryCatch(cor.test(rna_expression, protein_expression, 
                                                    method = "pearson")$p.value,
                                           error = function(cond){
                                             return(NA_real_) }),
                   spearman_rho = tryCatch(cor.test(rna_expression, protein_expression, 
                                                  method = "spearman", exact = T)$estimate,
                                         error = function(cond){
                                           return(NA_real_) }),
                   spearman_pval = tryCatch(cor.test(rna_expression, protein_expression, 
                                                     method = "spearman", exact = T)$p.value,
                                            error = function(cond){
                                              return(NA_real_) })) %>% 
  suppressWarnings() %>% 
  dplyr::mutate(pearson_padj = p.adjust(pearson_pval, method = "fdr"),
                spearman_padj = p.adjust(spearman_pval, method = "fdr")) %>% 
  ungroup() %>% 
  group_by(display_name) %>% 
  dplyr::filter(any(!is.na(pearson_r)) & any(!is.na(spearman_rho))) %>% 
  ungroup()
```
   
### Corplot   
```{r}
celltype_cor_mat = celltype_cors %>% 
  dplyr::mutate(spearman_rho = ifelse(is.na(spearman_rho),
                                   yes = 0,
                                   no = spearman_rho)) %>% 
  pivot_wider(id_cols = Celltype, names_from = display_name, values_from = spearman_rho) %>% 
  column_to_rownames("Celltype") %>% 
  data.matrix()
celltype_pval_mat = celltype_cors %>% 
  pivot_wider(id_cols = Celltype, names_from = display_name, values_from = spearman_padj) %>% 
  column_to_rownames("Celltype") %>% 
  data.matrix()

color_fun = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

celltype_cor_hm = 
  Heatmap(matrix = celltype_cor_mat,
        name = "Spearman rho",
        cluster_rows = T,
        cluster_columns = T,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_column_names = T,
        col = color_fun,
        cell_fun = function(j, i, x, y, w, h, fill) {
          if(is.na(celltype_pval_mat[i, j]))
          {
            grid.text("X", x, y)
          } else if(celltype_pval_mat[i, j] < 0.05){
            grid.text("*", x, y)
          }
        },
        column_names_rot = 45)

CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221117_celltype_corrplot.png",
    width = 16,
    height = 8, 
    units = "in",
    res = 300,
    family = "Arial")
celltype_cor_hm
dev.off()

celltype_cor_hm
```
   
## By cluster   
### Organize data   
```{r}
#includes non-COVID!
cluster_cor_data = read.csv("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_subcluster_expression_patient.csv") %>% 
  dplyr::filter(Cluster != "Mixed myeloid") %>% #meaningless
  left_join(., sample_conv) %>% 
  dplyr::select(-Patient) %>% 
  pivot_longer(cols = -c(Cluster, tc_pt_study_id),
               names_to = "gene",
               values_to = "log_rna_expression") %>% 
  dplyr::mutate(rna_expression = 10 ^ log_rna_expression) %>% 
  left_join(gene_conv) %>% 
  left_join(., complete_cytokines)
```   
   
### Run correlations   
```{r}
cluster_cors = cluster_cor_data %>% 
  group_by(display_name, Cluster) %>% 
  dplyr::summarize(pearson_r = tryCatch(cor.test(rna_expression, protein_expression,
                                                 method = "pearson")$estimate,
                                        error = function(cond){
                                          return(NA_real_) }),
                   pearson_pval = tryCatch(cor.test(rna_expression, protein_expression, 
                                                    method = "pearson")$p.value,
                                           error = function(cond){
                                             return(NA_real_) }),
                   spearman_rho = tryCatch(cor.test(rna_expression, protein_expression, 
                                                  method = "spearman", exact = T)$estimate,
                                         error = function(cond){
                                           return(NA_real_) }),
                   spearman_pval = tryCatch(cor.test(rna_expression, protein_expression, 
                                                     method = "spearman", exact = T)$p.value,
                                            error = function(cond){
                                              return(NA_real_) })) %>% 
  suppressWarnings() %>% 
  dplyr::mutate(pearson_padj = p.adjust(pearson_pval, method = "fdr"),
                spearman_padj = p.adjust(spearman_pval, method = "fdr")) %>% 
  ungroup() %>% 
  group_by(display_name) %>% 
  dplyr::filter(any(!is.na(pearson_r)) & any(!is.na(spearman_rho))) %>% 
  ungroup()
```
   
### Corplot   
```{r}
cluster_cor_mat = cluster_cors %>% 
  dplyr::mutate(spearman_rho = ifelse(is.na(spearman_rho),
                                   yes = 0,
                                   no = spearman_rho)) %>% 
  pivot_wider(id_cols = Cluster, names_from = display_name, values_from = spearman_rho) %>% 
  column_to_rownames("Cluster") %>% 
  data.matrix()
cluster_pval_mat = cluster_cors %>% 
  pivot_wider(id_cols = Cluster, names_from = display_name, values_from = spearman_padj) %>% 
  column_to_rownames("Cluster") %>% 
  data.matrix()

color_fun = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

cluster_cor_hm = 
  Heatmap(matrix = cluster_cor_mat,
        name = "Spearman rho",
        cluster_rows = T,
        cluster_columns = T,
        clustering_distance_rows = "euclidean",
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        show_column_names = T,
        col = color_fun,
        cell_fun = function(j, i, x, y, w, h, fill) {
          if(is.na(cluster_pval_mat[i, j]))
          {
            grid.text("X", x, y)
          } else if(cluster_pval_mat[i, j] < 0.05){
            grid.text("*", x, y)
          }
        },
        column_names_rot = 45)

CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/221117_cluster_corrplot.png",
    width = 16,
    height = 8, 
    units = "in",
    res = 300,
    family = "Arial")
cluster_cor_hm
dev.off()

cluster_cor_hm
```   
   
# Bar/dot plots with error bars, split by disease state   
### By cell type   
```{r}
protein_means = celltype_cor_data %>% 
  dplyr::filter(!is.na(diagnosis)) %>% 
  group_by(display_name) %>% 
  dplyr::mutate(mean_protein_covid = mean(subset(cur_data(), 
                                                 diagnosis == "COVID-19")$protein_expression, 
                                          na.rm = T)) %>% 
  ungroup() %>% 
  group_by(display_name, diagnosis, mean_protein_covid) %>% 
  dplyr::summarize(mean_protein = mean(protein_expression, na.rm = T)) %>% 
  ungroup()

celltype_contribution_scatterplot_patient = celltype_cor_data %>% 
  dplyr::filter(!is.na(diagnosis)) %>% 
  dplyr::mutate(is_moam = factor(Celltype == "MoAM")) %>% 
  group_by(display_name) %>% 
  dplyr::mutate(mean_protein_covid = mean(subset(cur_data(), 
                                                 diagnosis == "COVID-19")$protein_expression, 
                                          na.rm = T)) %>% 
  ungroup() %>% 
  ggplot(aes(x = reorder(display_name, -mean_protein_covid), y = rna_expression)) +
  facet_wrap(~diagnosis, ncol = 1) +
  geom_bar(data = protein_means, aes(y = mean_protein, fill = diagnosis), stat = "identity", alpha = 0.7) +
  stat_summary(aes(group = Celltype, shape = is_moam), geom = "point", fun = "mean") +
  stat_summary(aes(group = Celltype, label = Celltype), 
               geom = "text_repel", fun = "mean", min.segment.length = 0,
               size = 6, max.overlaps = 10) +
  theme_bw() +
  scale_fill_npg() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 24),
        axis.text.y = element_text(size = 24),
        axis.title.x = element_text(size = 36),
        axis.title.y = element_text(size = 36),
         strip.text = element_text(size = 24)) +
  labs(x = "Analyte", y = "Log-Normalized Expression") +
  scale_y_log10()

CairoPDF("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_celltype_contributions_scatter_patient.pdf",
    width = 12,
    height = 12, 
    family = "Arial")
celltype_contribution_scatterplot_patient
dev.off()
CairoPNG("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_celltype_contributions_scatter_patient.png",
    width = 12,
    height = 12, 
    units = "in",
    res = 300,
    family = "Arial")
celltype_contribution_scatterplot_patient
dev.off()

saveRDS(celltype_contribution_scatterplot_patient, "~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230125_celltype_contributions_scatter_patient.rds")

celltype_contribution_scatterplot_patient
```
      
# Are lung cytokine concentrations consistently higher than plasma?   
## Organize data   
```{r}
source_data_wide = readRDS("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230308_5PL_fit_data_complete.rds") %>% 
  dplyr::filter(diagnosis == "COVID-19") %>% 
  dplyr::select(tc_pt_study_id, study_id, sample_origin, 
                display_name, mean_concentration) %>% 
  dplyr::mutate(sample_origin = factor(case_when(sample_origin == "SERUM" ~ "Plasma",
                                                 sample_origin == "BAL" ~ "BAL"))) %>% 
  pivot_wider(names_from = sample_origin,
              values_from = mean_concentration) %>% 
  #keep only paired data
  dplyr::filter(!is.na(BAL) & !is.na(Plasma))

source_data_long = source_data_wide %>% 
  pivot_longer(cols = c(BAL, Plasma),
               names_to = "sample_origin",
               values_to = "mean_concentration")
```   
   
## Run paired stats
```{r}
paired_pvals = source_data_wide %>% 
  group_by(display_name) %>% 
  dplyr::summarize(pval = wilcox.test(x = cur_data()$BAL,
                                      y = cur_data()$Plasma,
                                      paired = TRUE)$p.value,
                   max_y = max(c(cur_data()$BAL, cur_data()$Plasma, na.rm = TRUE)),
                   log2FoldChange = log2(mean(cur_data()$BAL, na.rm = TRUE) /
                                           mean(cur_data()$Plasma, na.rm = TRUE))) %>% 
  dplyr::mutate(padj = p.adjust(pval, method = "fdr"),
                padj_formatted = format(padj, digits = 2, scientific = T),
                yval = log10(max_y) + 0.3) %>% 
  dplyr::filter(padj < 0.05) %>% 
  dplyr::arrange(desc(log2FoldChange))

write.csv(paired_pvals, "~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230701_BAL_over_plasma.csv")

paired_pvals_interest = paired_pvals %>% 
  dplyr::filter(display_name %in% c("CXCL10", "CCL8", "CCL2",
                                    "IL-6", "TNFa", "IL-1b"))
```
   
## Plot key cytokines   
```{r}
paired_source_plot = source_data_long %>% 
  dplyr::filter(display_name %in% c("CXCL10", "CCL8", "CCL2",
                                    "IL-6", "TNFa", "IL-1b")) %>% 
ggplot(aes(x = sample_origin, y = mean_concentration, fill = sample_origin)) +
  geom_point() +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~display_name, scales = "free_y") +
  geom_line(aes(group = study_id), linetype = 2) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 24),
        axis.text.y = element_text(size = 24),
        axis.title.x = element_text(size = 36),
        axis.title.y = element_text(size = 36),
         strip.text = element_text(size = 24)) +
  labs(x = "Sample Origin", y = "[Analyte] (pg/mL)") +
  scale_fill_manual(values = c(pal[4], pal[5])) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10), 
                     breaks = c(1e0, 1e1, 1e2, 1e3, 1e4, 1e5),
                     expand = expansion(mult = c(0, 0.15))) +
  geom_signif(data = paired_pvals_interest,
              aes(xmin = 1, xmax = 2, annotations = padj_formatted, y_position = yval),
              inherit.aes = F,
              tip_length = 0,
              manual = T, textsize = 9)

CairoPDF("~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230629_paired_source_plot.pdf",
    width = 12,
    height = 8, 
    family = "Arial")
paired_source_plot
dev.off()
saveRDS(paired_source_plot, "~/OneDrive - Northwestern University/Cytokine/Analysis/rgrant/230629_paired_source_plot.rds")

paired_source_plot
```